---
title: "Adjusted Values"
format:
  html:
    code-fold: TRUE
    toc: true
    toc-location: left
---

Index data pulled by David Merriman using Stata and FRED.

 * Gross Domestic Product: All Industry Total in Illinois (ILNGSP)
 * Consumer Price Index for All Urban Consumers: All Items in U.S. City Average (CPIAUCSL)
 * Producer Price Index by Commodity: Final Demand: Government Purchases, Nondefense (WPSFD49406)
 * Resident Population in Illinois (ILPOP)
 

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(cmapplot) # for recession line function

# values are in millions of dollars
totals <- readxl::read_xlsx("data/FY2025 Files/summary_file_FY25_PensionChange_2025_07_016.xlsx", sheet = "aggregated_totals_long")

# index data from FRED, pulled by DFM
adjustments <- readxl::read_xlsx("inputs/raw_index_data.xlsx")

totals2 <- totals |> left_join(adjustments, by = c("Year" = "year")) |>
  mutate(type = ifelse(type == "rev", "Revenue", "Expenditures"))

current_year = 2025

alea_theme <- function() {
  font <-"Whitney"
  
  ggplot2::theme(
    legend.position = "right",
    legend.title = element_blank(),
    
    panel.background = ggplot2::element_blank(),
    panel.grid.minor.x = ggplot2::element_blank(),
    panel.grid.major.y = element_line(color = "grey"),
    panel.grid.minor.y = element_line(color = "grey", 
                                      linetype = "dashed"),
    # panel.grid.major.x = ggplot2::element_blank(),
    axis.ticks = element_line(color = "gray"),
    axis.ticks.x = element_blank()
  )
  
}

theme_set(alea_theme())

# Custom billion format
label_billions <- function(digits = 1) {
  function(x) {
    number_format(accuracy = 10^-digits, suffix = "B")(
      x / 1e9
    )
  }
}

scale_y_billions <- function(..., digits = 1) {
  scale_y_continuous(labels = label_billions(digits), ...)
}


update_recessions <- function(url = NULL, quietly = FALSE){

  # Use default URL if user does not override
  if (is_null(url) | missing(url)) {
    url <- "https://data.nber.org/data/cycles/business_cycle_dates.json"
  }
  
   # locally bind variable names
  start_char <- end_char <- start_date <- end_date <- ongoing <- index <- peak <- trough <- NULL

  return(
    # attempt to download and format recessions table
    tryCatch({
      recessions <- jsonlite::fromJSON(url) %>%
        # drop first row trough
        dplyr::slice(-1) %>%
        # convert peaks and troughs...
        dplyr::mutate(
          # ...to R dates
          start_date = as.Date(peak),
          end_date = as.Date(trough),
          # ... and clean char strings
          start_char = format(start_date, "%b %Y"),
          end_char = format(end_date, "%b %Y")) %>%
        # confirm ascending and create row number
        dplyr::arrange(start_date) %>%
        mutate(index = row_number()) %>%
        mutate(
          # Flag unfinished recessions
          ongoing = case_when(
            is.na(end_date) & index == max(.$index) ~ T,
            TRUE ~ F),
          # set ongoing recession to arbitrary future date
          end_date = case_when(
            ongoing ~ as.Date("2200-01-01"),
            TRUE ~ end_date),
          # mark ongoing recession in char field
          end_char = case_when(
            ongoing ~ "Ongoing",
            TRUE ~ end_char)
          ) %>%
        # clean up
        select(start_char, end_char, start_date, end_date, ongoing)

      if (!quietly) {message("Successfully fetched from NBER")}

      # Return recessions
      recessions
    },
    error = function(cond){
      if (!quietly) message("WARNING: Fetch or processing failed. `NULL` returned.")
      return(NULL)
    }
    )
  )
}

#recess_table <- list(build_recessions(update))
recessions <- update_recessions()

```

# Inflated Values

Inflating makes old numbers “current” by pushing them up to today’s price level.


```{r warning = FALSE, message = FALSE}
#| layout-ncol: 2

ylabels <- c("Billions of Dollars",
             #"Billions of Dollars",
             "Billions of Dollars"
             )

vars <-  c("CPIAUCSL", "ILNGSP"#, "WPSFD49406"
           )

titles <- c("Adjusted by Inflation (CPIAUCSL)",
            "Adjusted by Illinois GDP (ILNGSP)"#,
         #   "Adjusted by Producer Price Index by Commodity (WPSFD4906)"
            )

for (i in seq_along(vars)) {
  
  var   <- vars[i]
  title <- titles[i]
    ylabel  <- ylabels[i]

    
    graph <- totals2 |> 
      mutate(
        base2024 = .data[[var]] / .data[[var]][Year == 2024],
        base1998 = .data[[var]] / .data[[var]][Year == 1998]
      )  |>
      
      mutate(adj_dollars = Dollars/1000 / base2024) |>  
      group_by(Year, type) |>
    summarize(adj_dollars = sum(adj_dollars, na.rm = TRUE), .groups = "drop") |>
    ggplot() + 
      geom_recessions(text = FALSE, update = recessions)+

      geom_line(aes(x = Year, y = adj_dollars, group = type, color = type), lwd = 1) +
      theme_minimal() +
    scale_y_continuous(labels=scales::dollar)+
      scale_color_manual(values = c("Revenue" = "black", "Expenditures" = "red")) + 

      scale_x_continuous(
        expand = c(0,0), 
        limits = c(1998, 2024), 
        breaks = c(1998, 2005, 2010, 2015, 2020, current_year-1)
      ) + 
      labs(
        title = title,
        y = ylabel,
        x = NULL
      ) + 
      theme(legend.title = element_blank()) 
  
  print(graph)
}


```

## Inflated Values per Person

```{r warning = FALSE, message = FALSE}
#| layout-ncol: 2

ylabels <- c("Dollars/Person",
             "Dollars/Person")

vars <-  c("CPIAUCSL", "ILNGSP")

titles <- c("Adjusted by Inflation (CPIAUCSL)",
            "Adjusted by Illinois GDP (ILNGSP)")

for (i in seq_along(vars)) {
  
  var   <- vars[i]
  title <- titles[i]
    ylabel  <- ylabels[i]

    
    graph <- totals2 |> 
      mutate(
        base2024 = .data[[var]] / .data[[var]][Year == 2024],
        base1998 = .data[[var]] / .data[[var]][Year == 1998]
      )  |>
      
      mutate(adj_dollars = Dollars*1000 / base2024) |>  
      group_by(Year, type) |>
    summarize(adj_dollars = sum(adj_dollars/ILPOP, na.rm = TRUE), .groups = "drop") |>
    ggplot() + 
      geom_recessions(text = FALSE, update = recessions)+

      geom_line(aes(x = Year, y = adj_dollars, group = type, color = type), lwd = 1) +
      theme_minimal() +
    scale_y_continuous(labels=scales::dollar)+
      scale_color_manual(values = c("Revenue" = "black", "Expenditures" = "red")) + 

      scale_x_continuous(
        expand = c(0,0), 
        limits = c(1998, 2024), 
        breaks = c(1998, 2005, 2010, 2015, 2020, current_year-1)
      ) + 
      labs(
        title = title,
        y = ylabel,
        x = NULL
      ) + 
      theme(legend.title = element_blank()) 
  
  print(graph)
}


```


# Deflated Values

Deflating is what economists do to get rid of inflation noise and see real growth, real price changes, or compare across time.


```{r warning = FALSE, message = FALSE}
#| layout-ncol: 2

ylabels <- c("Billions of Dollars",
             "Billions of Dollars")


vars <-  c("CPIAUCSL", "ILNGSP")

titles <- c("Adjusted by Inflation (CPIAUCSL)",
            "Adjusted by Illinois GDP (ILNGSP)")

for (i in seq_along(vars)) {
  
  var   <- vars[i]
  title <- titles[i]
  ylabel  <- ylabels[i]
  
  
  graph <- totals2 |> 
    mutate(
      base2024 = .data[[var]] / .data[[var]][Year == 2024],
      base1998 = .data[[var]] / .data[[var]][Year == 1998]
    )  |>
    
    mutate(adj_dollars = Dollars / base1998) |>  
    group_by(Year, type) |>
    summarize(adj_dollars = sum(adj_dollars, na.rm = TRUE), .groups = "drop") |>
    ggplot() + 
    geom_recessions(text = FALSE, update = recessions)+
    
    geom_line(aes(x = Year, y = adj_dollars, group = type, color = type), lwd = 1) +
    theme_minimal() +
    scale_y_continuous(labels=scales::dollar)+
    scale_color_manual(values = c("Revenue" = "black", "Expenditures" = "red")) + 
    
    scale_x_continuous(
      expand = c(0,0), 
      limits = c(1998, 2024), 
      breaks = c(1998, 2005, 2010, 2015, 2020, current_year-1)
    ) + 
    labs(
      title = title,
      y = ylabel,
      x = NULL
    ) + 
    theme(legend.title = element_blank()) 
  
  print(graph)
}


```

## Deflated Values per Person

```{r warning = FALSE, message = FALSE}
#| layout-ncol: 2

ylabels <- c("Dollars/Person",
             "Dollars/Person")

vars <-  c("CPIAUCSL", "ILNGSP")

titles <- c("Adjusted by Inflation (CPIAUCSL)",
            "Adjusted by Illinois GDP (ILNGSP)")

for (i in seq_along(vars)) {
  
  var   <- vars[i]
  title <- titles[i]
    ylabel  <- ylabels[i]

    
    graph <- totals2 |> 
      mutate(
        base2024 = .data[[var]] / .data[[var]][Year == 2024],
        base1998 = .data[[var]] / .data[[var]][Year == 1998]
      )  |>
      
      mutate(adj_dollars = Dollars*1000 / base1998) |>  
      group_by(Year, type) |>
    summarize(adj_dollars = sum(adj_dollars/ILPOP, na.rm = TRUE), .groups = "drop") |>
    ggplot() + 
      geom_recessions(text = FALSE, update = recessions)+

      geom_line(aes(x = Year, y = adj_dollars, group = type, color = type), lwd = 1) +
      theme_minimal() +
    scale_y_continuous(labels=scales::dollar)+
      scale_color_manual(values = c("Revenue" = "black", "Expenditures" = "red")) + 

      scale_x_continuous(
        expand = c(0,0), 
        limits = c(1998, 2024), 
        breaks = c(1998, 2005, 2010, 2015, 2020, current_year-1)
      ) + 
      labs(
        title = title,
        y = ylabel,
        x = NULL
      ) + 
      theme(legend.title = element_blank()) 
  
  print(graph)
}


```


