---
title: "Bond Proceeds and Debt - Sankey Graphs"
format: 
  html:
    code-fold: true
    df-print: paged
---

Bond proceeds were $2.7 billion in 2025.

Bond proceeds went to General Obligation BR&I, Build Illinois Bond, Transportation Bond Series A, Pension Acceleration Bond, Capital Development, Water Revolving, Transportation Bond Series B, Multimodal Transportation Bond, and Anti-Pollution Funds.



Debt Service Funds had $4.2 billion in expenditures in 2025.

However, only $1.96 is counted as "Debt Service" in Fiscal Futures categories. 


Capital Improvement expenditure items include expenditures from `agency=="511" | agency=="554" | agency=="574" | agency=="598" ~ "946"`


# Visualization of Bond Proceeds - Jahun's Trial

```{r}
library(tidyverse)
library(networkD3)
library(htmlwidgets)
```

```{r}

current_year <- 2025 # fiscal year, not calendar year
past_year=current_year-1

#rev_temp <- read_csv(paste0("data/FY", current_year, " Files/rev_temp.csv")) # |> filter(agency != 799)

#all_rev <- read_csv(paste0("data/FY", current_year, " Files/allrevfiles_2025.csv")) |> filter(fy > 2020)

all_rev <- read_csv("allrevfiles_2025.csv")  |> filter(fy > 2020)

rev_temp <- read_csv("rev_temp.csv") 



rev_bonds = rev_temp %>% 
  filter(fy == 2025) %>% 
  filter(rev_type == 72)

rev_bonds
```

```{r include = FALSE}
rev_bonds2 <- all_rev %>% 
  filter(fy == 2025) %>% 
  filter(source == "0571")

rev_bonds2
```


### 2 Levels (Fund -> Agency)

```{r}
#library(networkD3)
# 1) links: from (Source), to (Target), value 

link1 <- rev_bonds %>%
  mutate(
    source = paste0(fund_name_ab, " (Fund ", fund, ")"),
    target = paste0(agency_name, " (Agency ", agency, ")"),
    value  = receipts
  )

# Node List that contains the names for nodes
node1 <- data.frame(name = unique(c(link1$source, link1$target)),
                    stringsAsFactors = FALSE)

# Indexing: it starts from 0
link2 <- link1 %>% 
  mutate(
    source = match(source, node1$name) - 1,
    target = match(target, node1$name) - 1
  )

# 4) Adding Receipts to Labels
node_amounts <- link1 %>%
  select(source, target, value) %>%
  pivot_longer(cols = c(source, target),
               names_to = "type",
               values_to = "name") %>%
  group_by(name) %>%
  summarise(total_value = sum(value, na.rm = TRUE), .groups = "drop")

node1 <- node1 %>%
  left_join(node_amounts, by = "name")

node1 <- node1 %>%
  mutate(total_value = ifelse(is.na(total_value), 0, total_value))

node1 <- node1 %>%
  mutate(
    label = sprintf("%s | $%.1fM", name, total_value / 1e6)
  )

# 5) Making Sankey Diagram
p1 <- sankeyNetwork(
  Links   = link2,
  Nodes   = node1,
  Source  = "source",
  Target  = "target",
  Value   = "value",
  NodeID  = "label",  
  fontSize = 11,
  nodeWidth = 30,
width  = "100%"

)

p1
```


```{r}
library(ggalluvial)

# 1. Create a long-format dataset with alluvial structure
links_long <- rev_bonds |>
  mutate(
    fund_label   = paste0(fund_name_ab, " (Fund ", fund, ")"),
    agency_label = paste0(agency_name, " (Agency ", agency, ")")
  ) |>
  select(fund_label, fund, agency, agency_label, receipts)

# 2. Plot with ggalluvial using coord_flip()
ggplot(links_long,
       aes(axis1 = fund,
           axis2 = agency_label,
           y = receipts)) +
  geom_alluvium(aes(fill = fund), width = 1/12, alpha = 0.8, color = "grey40", reverse = FALSE) +
  geom_stratum(width = 1/10, reverse = FALSE, fill = "grey80") +   
  scale_fill_brewer(palette = "YlOrRd", direction = -1) +

  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3, reverse = FALSE) +
  scale_x_discrete(limits = c("Fund", "Agency"), expand = c(0.05, 0.05)) +
  coord_flip() +
  labs(
    y = "Receipts",
    x = NULL,
    title = "Fund → Agency Receipts (Top-to-Bottom Flow)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

```


```{r eval = FALSE}
#library(htmlwidgets)
# Adding Custom Labels
p1 <- onRender(
  p1,
  "
  function(el, x) {

    // Horizontal Spacing
    var offsetLeft  = -10;  // left-shift amount for the source label (px)
    var offsetRight =  40;  // right-shift amount for the target label (px)
    var dy          = 12;   // line spacing (px)

    d3.select(el).selectAll('.node text')
      .each(function(d) {

        var text  = d3.select(this);
        var full  = text.text();          
        var parts = full.split(' | ');    

        var xPos   = 0;
        var anchor = 'middle';

        // SOURCE Nodes
        if (d.sourceLinks.length > 0 && d.targetLinks.length == 0) {
          anchor = 'end';
          xPos   = offsetLeft;
        }
        // TARGET Nodes
        else if (d.targetLinks.length > 0 && d.sourceLinks.length == 0) {
          anchor = 'start';
          xPos   = offsetRight;
        }
        // Middle Nodes
        else {
          anchor = 'start';
          xPos   = offsetRight;
        }

        text.text(null);
        text.attr('text-anchor', anchor);

        for (var i = 0; i < parts.length; i++) {
          text.append('tspan')
              .attr('x', xPos)
              .attr('dy', i === 0 ? 0 : dy)
              .text(parts[i]);
        }
      });
  }
  "
)

p1



```


### 3 levels (Rev Source ->  Rev_type --> receipts ) for Total Amount


```{r}
# 1) links: from (Source), to (Target), value 
link3 <- rev_bonds %>%
  mutate(
    source = paste0(agency_name, " (Agency ", agency, ")"),
    target = rev_type_name,
    value  = receipts
  )

# Node List that contains the names for nodes
node2 <- data.frame(name = unique(c(link3$source, link3$target)),
                    stringsAsFactors = FALSE)

# 3) Indexing from 0
link4 <- link3 %>% 
  mutate(
    source = match(source, node2$name) - 1,
    target = match(target, node2$name) - 1
  )

# 4) Calculating Receipts and Adding to Label
node2_amounts <- link3 %>%
  select(source, target, value) %>%
  pivot_longer(cols = c(source, target),
               names_to = "type",
               values_to = "name") %>%
  group_by(name) %>%
  summarise(total_value = sum(value, na.rm = TRUE), .groups = "drop")

node2 <- node2 %>%
  left_join(node2_amounts, by = "name")

node2 <- node2 %>%
  mutate(total_value = ifelse(is.na(total_value), 0, total_value))

node2 <- node2 %>%
  mutate(
    label = sprintf("%s | $%.1fM", name, total_value / 1e6)
  )


# Merging Two Sections

links <- bind_rows(
  link1 %>% select(source, target, value),
  link3 %>% select(source, target, value)
)

nodes = bind_rows(node1, node2)
nodes = unique(nodes)

# Name index (from 0)
links <- links %>%
  mutate(
    source = match(source, nodes$name) - 1,
    target = match(target, nodes$name) - 1
  )



p2 <- sankeyNetwork(
  Links   = links,
  Nodes   = nodes,
  Source  = "source",
  Target  = "target",
  Value   = "value",
  NodeID  = "label",
  fontSize = 12,
  nodeWidth = 30,
  width  = "200%"
)

p2
```

```{r}
library(dplyr)
library(ggplot2)
library(ggalluvial)

# 1. Build long data with three layers:
# Fund  ->  Agency  ->  Revenue Type
sankey3 <- rev_bonds |>
  mutate(
    fund_label   = paste0(fund_name_ab, " (Fund ", fund, ")"),
    agency_label = paste0(agency_name, " (Agency ", agency, ")"),
    revtype_label = rev_type_name
  ) |>
  group_by(fund_label, agency_label, revtype_label) |>
  summarise(receipts = sum(receipts, na.rm = TRUE), .groups = "drop")

# 2. Three-axis alluvial with coord_flip() for top→bottom flow
ggplot(
  sankey3,
  aes(
    y     = receipts,
    axis1 = fund_label,
    axis2 = agency_label,
    axis3 = revtype_label
  )
) +
  geom_alluvium(aes(fill = fund_label), 
                width = 1/12, 
                alpha = 0.8, 
                color = "grey40",
                reverse = FALSE) +
  geom_stratum(width = 1/10, reverse = FALSE, fill = "grey80") +
    scale_fill_brewer(palette = "YlOrRd", direction = -1) +

  geom_text(
    stat  = "stratum",
    aes(label = after_stat(stratum)),
    size  = 3,
    reverse = FALSE
  ) +
  scale_x_discrete(
    limits = c("Fund", "Agency", "Revenue Type"),
    expand = c(0.05, 0.05)
  ) +
  coord_flip() +
  labs(
    y = "Receipts",
    x = NULL,
    title = "Revenue Types → Agencies → Funds"
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(angle = 0),
    axis.ticks   = element_blank(),
    panel.grid   = element_blank(),
    legend.position = "bottom"
  )

```

```{r}

# 1. Build long data with three layers:
# Fund  ->  Agency  ->  Revenue Type
sankey3 <- rev_bonds |>
  mutate(
    fund_label   = paste0(fund_name_ab, " (Fund ", fund, ")"),
    agency_label = paste0(agency_name, " (Agency ", agency, ")"),
    revtype_label = rev_type_name
  ) |>
  group_by(source_name_AWM, fund_label, agency_label, ) |>
  summarise(receipts = sum(receipts, na.rm = TRUE), .groups = "drop")

# 2. Three-axis alluvial with coord_flip() for top→bottom flow
ggplot(
  sankey3,
  aes(
    y     = receipts,
    axis1 = fund_label,
    axis2 = agency_label,
    axis3 = source_name_AWM
  )
) +
  geom_alluvium(aes(fill = fund_label), 
                width = 1/12, 
                alpha = 0.8, 
                color = "grey40",
                reverse = FALSE) +
  geom_stratum(width = 1/10, reverse = FALSE, fill = "grey80") +
  scale_fill_brewer(palette = "YlOrRd", direction = -1) +

  geom_text(
    stat  = "stratum",
    aes(label = after_stat(stratum)),
    size  = 3,
    reverse = FALSE
  ) +
  scale_x_discrete(
    limits = c("Fund", "Agency", "Source Type"),
    expand = c(0.05, 0.05)
  ) +
  coord_flip() +
  labs(
    y = "Receipts",
    x = NULL,
    title = "Funds → Agencies → Source Types"
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(angle = 0),
    axis.ticks   = element_blank(),
    panel.grid   = element_blank(),
    legend.position = "bottom"
  )

```


# Debt expenditures

```{r}

# exp_temp <- read_csv(paste0("data/FY", current_year, " Files/exp_temp.csv"))  |> filter(fy > 2020)
# all_exp <- read_csv(paste0("data/FY", current_year, " Files/allexpfiles_2025.csv"))  |> filter(fy > 2020)

all_exp <- read_csv("allexpfiles_2025.csv") 
exp_temp <- read_csv("exp_temp.csv")  |> filter(fy > 2020)


exp_transfers <- all_exp |> filter(fy == 2025 & agency == 799)
rev_transfers <- all_rev |> filter(fy ==2025 & agency == 799)

all_debt <- exp_temp %>%     # all principal, interest, and debt service except Tollway
  filter(fy == 2025) |>
    filter(#fund != "0455" &
             object %in% c(8811, 8813, 8800) ) #%>% 

 # filter(fund != "0455" & (object == "8811" | object == "8813" | object == "8800") )%>% 
  # group_by(fund, fund_name_ab, agency_name) %>% 
  # summarize(sum = sum(expenditure, na.rm=TRUE))

```


```{r debt-drop}
debt_drop <- exp_temp |> 
  filter(object == "8841" |  object == "8811")  
# escrow  OR  principle

# examine the debt costs we want to exclude
debt_drop |> filter(fy == 2025) |> arrange(desc(expenditure))

debt_drop |> group_by(fy) |> summarize(sum = sum(expenditure)) |> arrange(-fy)
```


```{r debt-keep}
debt_keep <- exp_temp |> 
  filter(fund != "0455" & (object == "8813" | object == "8800" )) 

# examine the debt costs we want to include
debt_keep |> filter(fy == 2025) |> arrange(desc(expenditure))

debt_keep |> group_by(fy) |> summarize(sum = sum(expenditure)) |> arrange(-fy)
```


```{r debt-items}
exp_temp <- anti_join(exp_temp, debt_drop) 
exp_temp <- anti_join(exp_temp, debt_keep)

debt_keep <- debt_keep |>
  mutate(
    agency = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(agency)),
    group = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(group)),
    in_ff = ifelse(group == "903", 1, as.character(in_ff)))

debt_keep_yearly <- debt_keep |> 
  group_by(fy, group) |> 
  summarize(debt_cost = sum(expenditure,na.rm=TRUE)/1000000) |> 
  select(-group)
```

```{r}

# 1. Build long data with three layers:
# Fund  ->  Agency  ->  Revenue Type
sankey3 <- all_debt |>
  mutate(
    fund_label   = paste0(fund_name_ab, " (Fund ", fund, ")"),
    agency_label = paste0(agency_name, " (Agency ", agency, ")"),
  ) |>
  group_by(fund, fund_label, agency_label, agency) |>
  summarise(expenditure = sum(expenditure, na.rm = TRUE), .groups = "drop")

# 2. Three-axis alluvial with coord_flip() for top→bottom flow
ggplot(
  sankey3,
  aes(
    y     = expenditure,
    axis1 = fund_label,
    axis2 = agency_label# ,
   # axis3 = source_name_AWM
  )
) +
  geom_alluvium(aes(fill = fund_label), 
                width = 1/12, 
                alpha = 0.8, 
                color = "grey40",
                reverse = FALSE) +
  geom_stratum(width = 1/10, reverse = FALSE, fill = "grey80") +
  scale_fill_brewer(palette = "YlOrRd", direction = -1) +

  geom_text(
    stat  = "stratum",
    aes(label = after_stat(stratum)),
    size  = 3,
    reverse = FALSE
  ) +
  scale_x_discrete(
    limits = c("Fund", "Agency"),
    expand = c(0.05, 0.05)
  ) +
  coord_flip() +
  labs(
    y = "Expenditure",
    x = NULL,
  #  title = "Funds → Agencies → Source Types"
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(angle = 0),
    axis.ticks   = element_blank(),
    panel.grid   = element_blank(),
    legend.position = "bottom"
  )

```


```{r}
capital <- exp_temp %>%     # all principal, interest, and debt service except Tollway
  filter(fy == 2025) |>
    filter(agency %in% c(511, 554, 574, 598) )

capital |> group_by(agency_name, fund_name_ab) |>
  summarize(sum = sum(expenditure))
```

```{r}
interfund_transfers <- exp_temp |> filter(object == "1993" & expenditure > 0) |> 
  mutate(expenditure = round(expenditure))

purchase_invests <- exp_temp|> filter(object == "1298" & expenditure > 0)
```

```{r}
#all_exp |> filter(fy > 2023 & agency == 799) |> write_csv("exp_transfers_2years.csv")

all_exp |> filter(fy>2023) |> write_csv("expenditures_2years.csv")
```

