```{r setup, warning = FALSE, message=FALSE, echo=FALSE}

library(tidyverse)
library(lubridate)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(data.table)
library(janitor)


knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```

::: {.callout-note appearance="minimal" icon="false"}
## Budget data with all observations for other researchers

This page describes the creation of `allexp_fy23_recoded_LASTDATEMODIFIED.csv` and `allrev_fy23_recoded_LASTDATEMODIFIED.csv`.
:::

This code is based on the (Everything.qmd) file used for the website and each years fiscal gap calculation, BUT it doesn't contain as many graphs or descriptions behind the code. It also has been modified slightly so that it DOES NOT drop observations that would normally be removed (e.g. principle payments, statutory transfers, etc.) during the data processing for the Fiscal Futures report. **It was created purely for the purpose of making a dataset that had all the observations received from the comptrollers office, with labels to go with the fund/agency/organization/revenuesource numbers, with non-duplicated fund numbers over time.**

This is a good faith effort to create a publicly available file that is labeled and coded nicely for other researchers. I cannot guarantee the accuracy but it should at least get you started on your task. I will try to do quality checks in the near future to make sure grouping variables are correct and that the binary variables created to indicate transfers, debt, etc. are also reliable. Use at your own risk and read the code closely! - AWM

::: {.callout-note appearance="minimal" icon="false"}
I tried to keep all observations, add additional variables not included in the raw data files, and have variable labels as clean as possible for any non-IGPA Fiscal Futures researchers that use the data for their own analyses.
:::

# Data Creation and Cleaning

See `Fiscal-Futures-Topics/code-cleaning.qmd`. 

::: {.callout-important appearance="simple" icon="false"}
Variable names must be identical to merge files together.
:::

Data files received include all observations through August 31 of each fiscal year. Data is requested and received after the lapse period closes on October 31.

Code below reads in the csv files in `Fiscal-Futures-Topics/code-cleaning.qmd`   (`rev_temp.csv` and `exp_temp.csv`). These files contain all years of data combined into one file. **It does have funds or agencies tracked consistently over time.**  Do not use this file for summing Fiscal Futures categories because it is just an intermediate step before recoding revenue and expenditure categories using Fiscal Futures methodology.  

{{< include ../../code-cleaning.qmd >}}


## Inspecting new FY22 data files

The numbers of funds, agencies, organizations, and revenue sources below were found by using pivot tables in the codebook files in Fiscal_Futures/data folder.

This can also be done using R and grouping new files by fund, agency, source number, source names, etc.. as is done in the code below.

Comparing all known agencies and funds from past years to the most recent year of data tells us if any new funds, agencies, sources, etc. began to be used or if an old fund number was reused for a new function. Tracking changes over time is a crucial step in the Fiscal Futures project methodology and beneficial to all research using Illinois state budget data.

> Disclaimer: The variable `local` existed when I began working on the project. I have not updated it or used it.

```{r}
## attempt at recreating items that were coded as amnesty==1 or local==1. 

## if similar to data that was inherited before joining the project, then I would apply it to FY19 to current FY to see if any revenue sources should have been coded as "local" or "amnesty"


local_rev_words <- c("MPEA", "MUNI", "COUNTY", "CNTY", "HOME RUL*", "NON HOME R*", "RTA", "LOCAL", "CHICAGO", "CITY", "R.O.T", "R O T", " ROT ")

amnesty_words <- c("AMNESTY")
```


> ::: {.callout-tip appearance="minimal" icon="false"}
> The [`funds_ab_in.xlsx`](inputs/funds_ab_in.xlsx) file contains all funds that have existed since 1998, if they still exist, indicates if fund numbers have been reused for different purposes, and is updated yearly with changes in fund numbers.
> :::


# Expenditure & Revenue Categorization

```{r}

read_csv(file = paste0("./data/FY", current_year, " Files/rev_temp.csv"))

read_csv(file = paste0("./data/FY", current_year, " Files/exp_temp.csv"))
```

## Modify Expenditure File

**Tax refunds**

Aggregate expenditures: Save tax refunds as negative revenue. Code refunds to match the rev_type codes (02=income taxes, 03 = corporate income taxes, 06=sales tax, 09=motor fuel tax, 24=insurance taxes and fees, 35 = all other tax refunds).


```{r tax-refunds}
tax_refund_long <- exp_temp %>%           # fund != "0401" # removes State Trust Funds
  filter(fund != "0401" & (object=="9910"|object=="9921"|object=="9923"|object=="9925")) %>%
  # keeps these objects which represent revenue, insurance, treasurer,and financial and professional reg tax refunds
  mutate(refund = case_when(
    fund=="0278" & sequence == "00" ~ "02", # for income tax refund
    fund=="0278" & sequence == "01" ~ "03", # tax administration and enforcement and tax operations become corporate income tax refund
     fund == "0278" & sequence == "02" ~ "02",
    object=="9921" ~ "21",                # inheritance tax and estate tax refund appropriation
    object=="9923" ~ "09",                # motor fuel tax refunds
    obj_seq_type == "99250055" ~ "06",    # sales tax refund
    fund=="0378" & object=="9925" ~ "24", # insurance privilege tax refund
    fund=="0001" & object=="9925" ~ "35", # all other taxes
      T ~ "CHECK"))                       # if none of the items above apply to the observations, then code them as CHECK 



# adds variable indicating if it was a refund or not to the exp_temp file #    
exp_temp <- left_join(exp_temp, tax_refund_long) %>%
  mutate(refund = ifelse(is.na(refund),"not refund", as.character(refund)))


# Normally these are dropped from expenditures for Fiscal Gap calculation
# exp_temp <- exp_temp %>% filter(refund == "Not refund")


```

**Create Pension Expenditure Identifier**


```{r}
knitr::read_chunk("path/to/_helpers.qmd")
```
```{r chunk-a, ref.label=c('chunk-c', 'chunk-b')}
# The code from chunk c and chunk b will appear here in the rendered output
```


See @sec-pensions for additional context.



Create indicator variable for if expenditure item is a pension expenditure. 

```{r pensions}

pension_check <- exp_temp %>% 
  
  mutate(
    pension = ifelse(object %in% c("4430","4431", "1159", "1160", 
                                   "1161", "1162", "1163","1164",
                                   "1165", "1166"), TRUE, FALSE),
  
    pension_label = case_when( 
    # object == "4430" & fund == "0825" ~ "Object 4430 - Pension Buyout/Benefits Paid Early",
    (object=="4430") ~ "Object 4430 - Benefits Paid to \nEmployees; EXCLUDED", # pensions, annuities, benefits
    (object=="4431") ~ "Object 4431 - State Contributions; INCLUDED", # 4431 = state payments into pension fund
    (obj_seq_type > "11590000" & obj_seq_type < "11660000")  ~ "Object 1160-1165 Employer Contributions \nto Pension Fund; INCLUDED",
    # objects 1159 to 1166 are all considered Retirement by Comptroller 
    
    TRUE ~ "0")) %>%  # All other observations coded as 0 for non-pension items
  
  # recodes specific instances of code anomalies from past years:
  mutate(pension = case_when(
    (object=="1298" & fund %in% pension_funds ) ~ "Object 1298 - Purchase of Investments; DROPPED", 
    # pension stabilization fund in 2022 
    # object == "1900" & fund == "0319" ~ "Fund 0319-Pension Stabilization", 
    object == "1900" & fund %in% pension_funds ~ "Fund 0319 - Pension Stabilization", 
    
    
    object == "4900" & fund %in% pension_funds ~ "Object 4900 - Awards/Grants; Weird 2010-2011 values",
    
    TRUE ~ as.character(pension)) ) %>% 
  filter(pension != "0" )

```

```{r pensions-POB}

# special accounting of pension obligation bond (POB)-funded contributions to JRS, SERS, GARS, TRS 

exp_temp <- exp_temp %>% 
  # change object for 2010 and 2011, retirement expenditures were bond proceeds and would have been excluded
  mutate(object = ifelse((pension >0 & in_ff == "0"), "4431", object)) %>% 
  # changes weird teacher & judge retirement system  pensions object to normal pension object 4431
  mutate(pension =  ifelse(pension >0 & in_ff == "0", 6, pension)) %>% # coded as 6 if it was supposed to be excluded. 
  mutate(in_ff = ifelse(pension>0, "1", in_ff))

table(exp_temp$pension) 


# all other pensions objects  codes get agency code 901 for State Pension Contributions
exp_temp <- exp_temp %>% 
  mutate(agency = ifelse(pension>0, "901", as.character(agency)),
         agency_name = ifelse(agency == "901", "State Pension Contributions", as.character(agency_name)))


```

**Drop Interfund transfers**

These items are dropped in the fiscal futures model. They have been left in for other researchers to use in the rev_temp and exp_temp files in the data folder.

-   object == 1993 is for interfund cash transfers
-   agency == 799 is for statutory transfers
-   object == 1298 is for purchase of investments and is not spending EXCEPT for costs in 2010 and 2011 (and were recoded already to object == "4431"). Over 168,000 observations remain.

```{r transfers}

#### these observations are NOT dropped in data files available to other researchers
# transfers_drop <- exp_temp %>% filter(
#   agency == "799" | # statutory transfers
#            object == "1993" |  # interfund cash transfers
#            object == "1298") # purchase of investments
# 
# exp_temp <- anti_join(exp_temp, transfers_drop)


exp_temp <- exp_temp %>% 
  mutate(drop_reason = ifelse(
    agency == "799" |             # statutory transfers
      object == "1993" |  # interfund cash transfers
      object == "1298",  # purchase of investments
    "Transfers", drop_reason) )

```

**State employee healthcare costs**

Relevant funds:

-   Fund = 0457 is "Group insurance premium", in_ff = 1
-   Fund = 0193 is "Local govt health insurance reserve", in=ff = 0
-   fund = 0477 is "Community College Health Insurance", in=ff = 0.
    -   had large amount in early years
-   Fund = 0907 = health insurance reserve, in_ff = 1
-   Fund = 9939 is "group self-insurers' insolv", in_ff = 1
-   Fund = 0940 is Self-Insurers security, in_ff = 0
-   Fund = 0739 is Group Workers Comp Pool Insol, in_ff = 1
-   eehc = 0 means it is NOT a state healthcare cost but it is an employer contribution of some type to some fund
-   eehc = 1 means it is a state employee healthcare cost and it is an employer contribution to health insurance

If observation is a group insurance contribution, then the expenditure amount is set to \$0 (essentially dropped from analysis).

::: {.callout-note appearance="minimal" icon="false"}
I code group insurance observations as `drop_reason == "Group Insurance Contribution` in the code chunk below. Otherwise expenditure values would be changed to 0 for some observations that other researchers may want to use.
:::

```{r state-employee-healthcare}

exp_temp <- exp_temp %>% 
  mutate(drop_reason = ifelse(
    # group insurance contributions for 1998-2005 and 2013-present
  (fund == "0001" & (object == "1180" | object =="1900") & agency == "416" & appr_org=="20") |
   
     # group insurance contributions for 2006-2012
    (fund == "0001" & object == "1180" & agency == "478" & appr_org=="80") |
  
  # group insurance contributions from road fund
  # coded with 1900 for some reason??
  (fund == "0011" & object == "1900" & agency == "416" & appr_org=="20"), 
  "Group Insurance Contribution", drop_reason) )
  
  
# State employee healthcare indicator   
exp_temp <- exp_temp |>
 # turns specific items into State Employee Healthcare 
  mutate(fiscal_future_cat = 
    ifelse(
       # central management Bureau of benefits using health insurance reserve 
      (fund=="0907" & (agency=="416" & appr_org=="20")) |  
         
      # agency = 478: healthcare & family services using health insurance reserve - ioc stopped using this in 2012
     ( fund=="0907" & (agency=="478" & appr_org=="80") ), "State Employee Healthcare", fiscal_future_cat),
    
    # still assign fiscal future category like we normally do in the group variable
    # defaults to agency code until manually recoded
    group = ifelse(fiscal_future_cat == "State Employee Healthcare", "904", as.character(agency)) )




```

**Local Transfers**

Separate transfers to local from parent agencies that come from DOR(492) or Transportation (494). Treats muni revenue transfers as expenditures, not negative revenue.

The six corresponding revenue items are:

-   Local share of Personal Income Tax
    -   Individual Income Tax Pass-Through New 2021 (source 2582).
-   Local share of General Sales Tax
-   Personal Property Replacement Tax on Business Income
-   Personal Property Replacement Tax on Public Utilities
-   Local share of Motor Fuel Tax - Transportation Renewal Fund 0952

```{r transfers-to-local}
exp_temp <- exp_temp %>% 
  mutate(
  agency = case_when(fund=="0515" & object=="4470" & type=="08" ~ "971", # income tax to local governments
                     fund=="0515" & object=="4491" & type=="08" & sequence=="00" ~ "971", # object is shared revenue payments
                     fund=="0802" & object=="4491" ~ "972", #pprt transfer
                     fund=="0515" & object=="4491" & type=="08" & sequence=="01" ~ "976", #gst to local
                     fund=="0627" & object=="4472"~ "976" , # public transportation fund but no observations exist
                     fund=="0648" & object=="4472" ~ "976", # downstate public transportation, but doesn't exist
                     fund=="0515" & object=="4470" & type=="00" ~ "976", # object 4470 is grants to local governments
                    object=="4491" & (fund=="0188"|fund=="0189") ~ "976",
                     fund=="0187" & object=="4470" ~ "976",
                     fund=="0186" & object=="4470" ~ "976",
                    object=="4491" & (fund=="0413"|fund=="0414"|fund=="0415")  ~ "975", #mft to local
                  fund == "0952"~ "975", # Added Sept 29 2022 AWM. Transportation Renewal MFT
                    TRUE ~ as.character(agency)),
  
  agency_name = case_when(agency == "971"~ "INCOME TAX 1/10 TO LOCAL",
                          agency == "972" ~ "PPRT TRANSFER TO LOCAL",
                          agency == "975" ~ "MFT TO LOCAL",
                          agency == "976" ~ "GST TO LOCAL",
                          TRUE~as.character(agency_name)),
  group = ifelse(agency>"970" & agency < "977", as.character(agency), as.character(group)))

```

```{r drop-local-transfers}
#transfers_long <- exp_temp %>% 
#  filter((group == "971" |group == "972" | group == "975" | group == "976"))# | fund == "0325")


### would drop local transfer items
### commented out so transfers are kept in dataframe

# exp_temp <- anti_join(exp_temp, transfers_long)


```

```{r, drop-inffis0}

#### drops all other in_ff = 0 funds  ####

#exp_temp <- exp_temp %>% filter(in_ff == 1) 

```

**Debt Service**

Debt Service expenditures include interest payment on both short-term and long-term debt. We do not include escrow or principal payments.

::: {.callout-note appearance="minimal" icon="false"}
Commented out both of these lines in chunk below: `exp_temp <- anti_join(exp_temp, debt_drop)` and `exp_temp <- anti_join(exp_temp, debt_keep)`
:::

```{r debt-service}
debt_drop <- exp_temp %>% filter(object == "8841" |  object == "8811")
# escrow  OR  principle



debt_keep <- exp_temp %>% 
  filter(fund != "0455" & (object == "8813" | object == "8800" ))  %>% 
  mutate(debt_keep = 1)


# exp_temp <- anti_join(exp_temp, debt_keep)

#changed from debt_keep to exp_temp for recoding and keeping obs for other researchers.
exp_temp <- exp_temp %>%
  mutate(
    agency = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(agency)),
    group = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(group)),
    in_ff = ifelse(group == "903", 1, in_ff ))

# exp_temp <- anti_join(exp_temp, debt_drop) 


```

**Medicaid**

That portion of the Healthcare and Family Services (or Public Aid in earlier years, agency code 478) budget for Medical (appr_organization code 65) for awards and grants (object codes 4400 and 4900).

-   Department of Healthcare and Family Services administers medical assistance programs (e.g. Medicaid, CHIP, All Kids). The Fiscal Futures model moves Medicaid and CHIP expenditures to their own expenditure category. State funded programs remain in DHFS.

    Side note: State CURE healthcare provider relief will remain in the Medicaid expenditure category due to the nature of it being federal funds providing public health services and funding to locations that provide public services.

-   Uses same appropriation name of "HEALTHCARE PROVIDER RELIEF" and fund == 0793 and obj_seq_type == 49000000. So can defend including healthcare provider relief as Medicaid expenditure.

-   Done in next section

### Add Other Fiscal Future group codes

```{r group-codes}


exp_temp <- exp_temp %>%
  #mutate(agency = as.numeric(agency) ) %>%
  # arrange(agency)%>%
  mutate(
    group = case_when(
      agency>"100"& agency<"200" ~ "910", # legislative
      
      agency == "528"  | (agency>"200" & agency<"300") ~ "920", # judicial
      pension>0  ~ "901", # pensions
      (agency>"309" & agency<"400") ~ "930",    # elected officers
      
      agency == "586" ~ "959", # create new K-12 group

      agency=="402" | agency=="418" | agency=="478" | agency=="444" | agency=="482" ~ as.character(agency), # aging, Children and Family Services (DCFS), Healthcare and Family Services (DHFS), human services (DHS), public health (DPH)
      T ~ as.character(group))
    ) %>%      

  
  mutate(group = case_when(
    agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400") ~ "945", # separates CHIP from health and human services and saves it as Medicaid
    
    agency == "586" & fund == "0355" ~ "945",  # 586 (Board of Edu) has special education which is part of medicaid
    
    # OLD CODE: agency == "586" & appr_org == "18" ~ "945", # Spec. Edu Medicaid Matching
    
    agency=="425" | agency=="466" | agency=="546" | agency=="569" | agency=="578" | agency=="583" | agency=="591" | agency=="592" | agency=="493" | agency=="588" ~ "941", # public safety & Corrections
    
     agency=="406" | agency=="420" | agency=="494" |  agency=="557" ~ as.character(agency), # econ devt & infra, tollway
    
    agency=="511" | agency=="554" | agency=="574" | agency=="598" ~ "946",  # Capital improvement
    
    agency=="422" | agency=="532" ~ as.character(agency), # environment & nat. resources
    
    agency=="440" | agency=="446" | agency=="524" | agency=="563"  ~ "944", # business regulation
    
    agency=="492" ~ "492", # revenue
    
    agency == "416" ~ "416", # central management services
    agency=="448" & fy > 2016 ~ "416", #add DoIT to central management 
    
    T ~ as.character(group))) %>%
  
  
  mutate(group = case_when(
    # agency=="684" | agency=="691"  ~ as.character(agency), # moved under higher education in next line. 11/28/2022 AWM
    
    agency=="692" | agency=="695" | agency == "684" |agency == "691" | (agency>"599" & agency<"677") ~ "960", # higher education
    
    agency=="427"  ~ as.character(agency), # employment security
    
    agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments
    
    # other boards & Commissions
    agency=="503" | agency=="509" | agency=="510" | agency=="565" |agency=="517" | agency=="525" | agency=="526" | agency=="529" | agency=="537" | agency=="541" | agency=="542" | agency=="548" |  agency=="555" | agency=="558" | agency=="559" | agency=="562" | agency=="564" | agency=="568" | agency=="579" | agency=="580" | agency=="587" | agency=="590" | agency=="527" | agency=="585" | agency=="567" | agency=="571" | agency=="575" | agency=="540" | agency=="576" | agency=="564" | agency=="534" | agency=="520" | agency=="506" | agency == "533" ~ "949", 
    
    # non-pension expenditures of retirement funds moved to "Other Departments"
    # should have removed pension expenditures already from exp_temp in Pensions step above
    agency=="131" | agency=="275" | agency=="589" |agency=="593"|agency=="594"|agency=="693" ~ "948",
    
    T ~ as.character(group))) %>%

  mutate(group_name = 
           case_when(
             group == "416" ~ "Central Management",
             group == "478" ~ "Healthcare and Family Services",
             group == "482" ~ "Public Health",
             group == "900" ~ "NOT IN FRAME",
             group == "901" ~ "STATE PENSION CONTRIBUTION",
             group == "903" ~ "DEBT SERVICE",
             group == "910" ~ "LEGISLATIVE"  ,
             group == "920" ~ "JUDICIAL" ,
             group == "930" ~ "ELECTED OFFICERS" , 
             group == "940" ~ "OTHER HEALTH-RELATED", 
             group == "941" ~ "PUBLIC SAFETY" ,
             group == "942" ~ "ECON DEVT & INFRASTRUCTURE" ,
             group == "943" ~ "CENTRAL SERVICES",
             group == "944" ~ "BUS & PROFESSION REGULATION" ,
             group == "945" ~ "MEDICAID" ,
             group == "946" ~ "CAPITAL IMPROVEMENT" , 
             group == "948" ~ "OTHER DEPARTMENTS" ,
             group == "949" ~ "OTHER BOARDS & COMMISSIONS" ,
             group == "959" ~ "K-12 EDUCATION" ,
             group == "960" ~ "UNIVERSITY EDUCATION" ,
             group == agency ~ as.character(group),
             TRUE ~ "Check name"),
         year = fy)

exp_temp %>% filter(group_name == "Check name")

#write_csv(exp_temp, "all_expenditures_recoded.csv")

```

All expenditures recoded but not aggregated: Allows for inspection of individual expenditures within larger categories. This stage of the data is extremely useful for any researcher interested in Illinois fiscal health. This version of data produced by this file still includes observations that would normally be dropped.

This was created in an attempt to help other researchers have the most complete possible data set. Output was not used for Fiscal Futures documents. We cannot guarantee that output is 100% correct due to changes made in code to keep all observations compared to the `Everything.qmd` file.

## Modify Revenue data

Revenue Categories NOT included in Fiscal Futures:\
- 32. Garnishment-Levies. (State is fiduciary, not beneficiary.)\
- 45. Student Fees-Universities. (Excluded from state-level budget.)\
- 51. Retirement Contributions (of individuals and non-state entities).\
- 66. Proceeds, Investment Maturities. (Not sustainable flow.)\
- 72. Bond Issue Proceeds. (Not sustainable flow.)\
- 75. Inter-Agency Receipts.\
- 79. Cook County Intergovernmental Transfers. (State is not beneficiary.)\
- 98. Prior Year Refunds.\
- 99. Statutory Transfers.

**All Other Sources**

Expanded to include the following smaller sources:\
- 30. Horse Racing Taxes & Fees.\
- 60. Other Grants and Contracts.\
- 63. Investment Income.

For aggregating revenue, use the rev_1998_2022 dataframe, join the funds_ab_in_2022 file to it, and then join the ioc_source_type file to the dataset. Remember: You need to update the funds_ab_in and ioc_source_type file every year!

```{r, warning = FALSE, message=FALSE}
# fund info to revenue for all years
rev_temp <- inner_join(rev_1998_2023, funds_ab_in_2023, by = "fund") %>% arrange(source)

# need to update the ioc_source_type file every year! 
ioc_source_type <- readxl::read_xlsx("ioc_source_updated23_AWM.xlsx")

rev_temp <- left_join(rev_temp, ioc_source_type, by = "source")
# automatically used source, source name does not match for the join to work using source_name



# recodes old agency numbers to consistent agency number
rev_temp <- rev_temp %>% 
  mutate(agency = case_when(
    (agency=="438"| agency=="475" |agency == "505") ~ "440",
    # financial institution &  professional regulation &
     # banks and real estate  --> coded as  financial and professional reg
    agency == "473" ~ "588", # nuclear safety moved into IEMA
    (agency =="531" | agency =="577") ~ "532", # coded as EPA
    (agency =="556" | agency == "538") ~ "406", # coded as agriculture
    agency == "560" ~ "592", # IL finance authority (fire trucks and agriculture stuff)to state fire marshal
    agency == "570" & fund == "0011" ~ "494",   # city of Chicago road fund to transportation
    TRUE ~ (as.character(agency)))) 
```

#### Federal Revenue

The Fiscal Futures Model divides federal funds (IOC revenue type = 57) into Medicaid, Transportation, and All Other Federal Funds. These can include: Health and Human Services Grants, Federal Stimulus Package, Department of Education Grants, Department of Transportation Grants, Department of Agriculture Grants, TANF Grants, and Department of Labor Grants.

**Federal Medicaid:** DHFS receives money for Medicaid that is deposited into the General Revenue Fund. There are also Special State Funds used for Medicaid that receive specific revenues (e.g. Healthcare Provider Taxes) which are matched with Federal Funds. The federal receipts in these special funds are aggregated and added to the federal receipts in the GRF that are received by DHFS.

Sources:

-   618 = Health and Human Services (not used)
-   660 = HHS/Hospital Participation
-   676 = Medical Assistance
-   692 = Medical Assistance
-   1552 = DHHS/ FFP-Medicaid Rehab Option
-   2306 = Enhanced Fed Fin PART-ARRA
-   2076 = IDPH-HHS/CMS
-   2364 = Department of Insurance (not used)

Other potential medicaid sources from CTRL-Fing "med" in revenue sources:\
Sources: 2104 = Medicare Part D & 675 = Medical Administration

**Federal Transportation:** If Agency is 494 and considered Federal Revenue, then it is recoded to its own category of "Federal Transportation".

```{r create-rev-federal-transfers}
#rev_temp <- rev_temp %>% filter(in_ff==1)

rev_temp <- rev_temp %>% 
  mutate(
    rev_type = ifelse(rev_type=="57" & agency=="478" & (source=="0618"|source=="2364"|source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692"), "58", rev_type),
    rev_type_name = ifelse(rev_type=="58", "Federal Medicaid Reimbursements", rev_type_name),
    rev_type = ifelse(rev_type=="57" & agency=="494", "59", rev_type),
    rev_type_name = ifelse(rev_type=="59", "Federal Transportation", rev_type_name),
    rev_type_name = ifelse(rev_type=="57", "Federal - Other", rev_type_name),
    rev_type = ifelse(rev_type=="6", "06", rev_type),
    rev_type = ifelse(rev_type=="9", "09", rev_type)) 

```

```{r, code-fed-stimulus-source}
# only labels observations where the source was specifically named "federal stimulous package"
# does not code other COVID legislative funds that went under other names and had more specific purposes.
rev_temp <- rev_temp %>% 
  mutate(covid_dollars = ifelse(source_name_AWM == "FEDERAL STIMULUS PACKAGE",1,0))
```

#### Health Insurance Premiums from Employees

Insurance premiums for employees is coded below but it is NOT used in the fiscal futures model. Employee and employer premiums are considered rev_51 and dropped from analysis in later step.

0120 = ins prem-option life\
0120 = ins prem-optional life/univ

0347 = optional health - HMO\
0348 = optional health - dental\
0349 = optional health - univ/local SI\
0350 = optional health - univ/local\
0351 = optional health - retirement\
0352 = optional health - retirement SI\
0353 = optional health - retire/dental\
0354 = optional health - retirement hmo

2199-2209 = various HMOs, dental, health plans from Health Insurance Reserve (fund)

```{r insurance-premiums}

#collect optional insurance premiums to fund 0907 for use in eehc expenditure  
rev_temp <- rev_temp %>% 
  mutate(
    #variable not used in aggregates, but could be interesting for other purposes
    employee_premiums = ifelse(fund=="0907" & (source=="0120"| source=="0121"| (source>"0345" & source<"0357")|(source>"2199" & source<"2209")), 1, 0),
    
    # adds more rev_type codes
    rev_type = case_when(
      fund =="0427" ~ "12", # pub utility tax
      fund == "0742" | fund == "0473" ~ "24", # insurance and fees
      fund == "0976" ~ "36",# receipts from rev producing
      fund == "0392" |fund == "0723" ~ "39", # licenses and fees
      fund == "0656" ~ "78", #all other rev sources
      TRUE ~ as.character(rev_type)))
# if not mentioned, then rev_type as it was



# # optional insurance premiums = employee insurance premiums

# emp_premium <- rev_temp %>%
#   group_by(fy, employee_premiums) %>%
#   summarize(employee_premiums_sum = sum(receipts)/1000000) %>%
#   filter(employee_premiums == 1) %>%
#   rename(year = fy) %>% 
#   select(-employee_premiums)

# emp_premium_long <- rev_temp %>%  filter(employee_premiums == 1)
# 381 observations have employee premiums == 1


# drops employee premiums from revenue
# rev_temp <- rev_temp %>% filter(employee_premiums != 1)
# should be dropped in next step since rev_type = 51


```

Employee premiums are dropped in the following steps. In FY21, employee premiums were subtracted from state healthcare costs on the expenditure side to calculate a "Net Healthcare Cost" but that methodology has been discontinued. Totals were practically unchanged: revenue from employee premiums is also very small.

### Transfers in and Out:

Funds that hold and disperse local taxes or fees are dropped from the analysis. Then other excluded revenue types are also dropped.

Drops Blank, Student Fees, Retirement contributions, proceeds/investments, bond issue proceeds, interagency receipts, cook IGT, Prior year refunds.

```{r}
### Commented out so that revenue source observations are not dropped.
## normally would remove all observations where in_ff was 0

# rev_temp <- rev_temp %>% 
#   filter(in_ff == 1) %>% 
#   mutate(local = ifelse(is.na(local), 0, local)) %>% # drops all revenue observations that were coded as "local == 1"
#   filter(local != 1)
```

```{r}

### code chunk not used.

# 1175 doesnt exist?
in_from_out <- c("0847", "0867", "1175", "1176", "1177", "1178", "1181", "1182", "1582", "1592", "1745", "1982", "2174", "2264")

# what does this actually include:
# all are items with rev_type = 75 originally. 
in_out_df <- rev_temp %>%
  mutate(infromout = ifelse(source %in% in_from_out, 1, 0)) %>%
  filter(infromout == 1)

rev_temp <- rev_temp %>% 
  mutate(rev_type_new = ifelse(source %in% in_from_out, "76", rev_type))
# if source contains any of the codes in in_from_out, code them as 76 (all other rev).
# I end up excluding rev_76 in later steps
```

Commented out drop_type code in chunk below to avoid dropping these observations for other researchers:

```{r droprevtypes}
# # revenue types to drop
# drop_type <- c("32", "45", "51", 
#                "66", "72", "75", "79", "98")
# 
# # drops Blank, Student Fees, Retirement contributions, proceeds/investments,
# # bond issue proceeds, interagency receipts, cook IGT, Prior year refunds.
# 
# 
# rev_temp <- rev_temp %>% filter(!rev_type_new %in% drop_type)
# # keep observations that do not have a revenue type mentioned in drop_type



# combines smallest 4  categories to to "Other"
# they were the 4 smallest in past years, are they still the 4 smallest? 

rev_temp <- rev_temp %>%  
 mutate(rev_type_new = ifelse(rev_type=="30" | rev_type=="60" | rev_type=="63" | rev_type=="76", "78", rev_type_new))


#table(rev_temp$rev_type_new)  # check work



rm(rev_1998_2022)
rm(exp_1998_2022)


#write.csv(exp_temp, "allexp_fy22_recoded_02242023.csv")
#write.csv(rev_temp, "allrev_fy22_recoded_02242023.csv")

write.csv(exp_temp, "allexp_fy23_recoded_Dec2023.csv")
write.csv(rev_temp, "allrev_fy23_recoded_Dec2023.csv")

```

```{r final-ffrev-table, eval = FALSE}
ff_rev <- rev_temp %>% 
  group_by(rev_type_new, fy) %>% 
  summarize(sum_receipts = sum(receipts, na.rm=TRUE)/1000000 ) %>%
  pivot_wider(names_from = "rev_type_new", values_from = "sum_receipts", names_prefix = "rev_")

ff_rev<- left_join(ff_rev, tax_refund)

#ff_rev <- left_join(ff_rev, pension2_fy22, by=c("fy" = "year"))

#ff_rev <- left_join(ff_rev, eehc2_amt) 
ff_rev <- mutate_all(ff_rev, ~replace_na(.,0))


ff_rev <- ff_rev %>%
  mutate(rev_02 = rev_02 - ref_02,
         rev_03 = rev_03 - ref_03,
         rev_06 = rev_06 - ref_06,
         rev_09 = rev_09 - ref_09,
         rev_21 = rev_21 - ref_21,
         rev_24 = rev_24 - ref_24,
         rev_35 = rev_35 - ref_35

      #   rev_78new = rev_78 #+ pension_amt #+ eehc
         ) %>% 
  select(-c(ref_02:ref_35, rev_99, rev_NA, rev_76#, pension_amt , rev_76,
          #  , eehc
            ))

ff_rev


```

```{r eval=FALSE}

## Since I already pivot_wider()ed the table in the previous code chunk, I now change each column's name by using rename() to set new variable names. Ideally the final dataframe would have both the variable name and the variable label but I have not done that yet.


aggregate_rev_labels <- ff_rev %>%
  rename("INDIVIDUAL INCOME TAXES, gross of local, net of refunds" = rev_02,
         "CORPORATE INCOME TAXES, gross of PPRT, net of refunds" = rev_03,
         "SALES TAXES, gross of local share" = rev_06 ,
         "MOTOR FUEL TAX, gross of local share, net of refunds" = rev_09 ,
         "PUBLIC UTILITY TAXES, gross of PPRT" = rev_12,
         "CIGARETTE TAXES" = rev_15 ,
         "LIQUOR GALLONAGE TAXES" = rev_18,
         "INHERITANCE TAX" = rev_21,
         "INSURANCE TAXES&FEES&LICENSES, net of refunds" = rev_24 ,
         "CORP FRANCHISE TAXES & FEES" = rev_27,
       # "HORSE RACING TAXES & FEES" = rev_30,  # in Other
         "MEDICAL PROVIDER ASSESSMENTS" = rev_31 ,
         # "GARNISHMENT-LEVIES " = rev_32 , # dropped
         "LOTTERY RECEIPTS" = rev_33 ,
         "OTHER TAXES" = rev_35,
         "RECEIPTS FROM REVENUE PRODUCNG" = rev_36, 
         "LICENSES, FEES & REGISTRATIONS" = rev_39 ,
         "MOTOR VEHICLE AND OPERATORS" = rev_42 ,
         #  "STUDENT FEES-UNIVERSITIES" = rev_45,   # dropped
         "RIVERBOAT WAGERING TAXES" = rev_48 ,
         # "RETIREMENT CONTRIBUTIONS " = rev_51, # dropped
         "GIFTS AND BEQUESTS" = rev_54, 
         "FEDERAL OTHER" = rev_57 ,
         "FEDERAL MEDICAID" = rev_58, 
         "FEDERAL TRANSPORTATION" = rev_59 ,
         #"OTHER GRANTS AND CONTRACTS" = rev_60, #other
       # "INVESTMENT INCOME" = rev_63, # other
         # "PROCEEDS,INVESTMENT MATURITIES" = rev_66 , #dropped
         # "BOND ISSUE PROCEEDS" = rev_72,  #dropped
         # "INTER-AGENCY RECEIPTS" = rev_75,  #dropped
      #  "TRANSFER IN FROM OUT FUNDS" = rev_76,  #other
         "ALL OTHER SOURCES" = rev_78,
         # "COOK COUNTY IGT" = rev_79, #dropped
         # "PRIOR YEAR REFUNDS" = rev_98 #dropped
  ) 

aggregate_rev_labels

```

### 

Create exp_970 for all local government transfers (exp_971 + exp_972 + exp_975 + exp_976).

```{r, eval = FALSE}
### EVAL = FALSE in code chunk

ff_exp <- exp_temp %>% 
  group_by(fy, group) %>% 
  summarize(sum_expenditures = sum(expenditure, na.rm=TRUE)/1000000 ) %>%
  pivot_wider(names_from = "group", values_from = "sum_expenditures", names_prefix = "exp_")%>%
  
    left_join(debt_keep_yearly) %>%
  mutate(exp_903 = debt_cost) %>%


  # join local transfers and create exp_970
  left_join(transfers) %>%
  mutate(exp_970 = exp_971 + exp_972  + exp_975 + exp_976)

ff_exp<- ff_exp %>% select(-c(debt_cost, exp_971:exp_976)) # drop unwanted columns
ff_exp
```

> To do: add group labels to go with group numbers. That was done after aggregating and pivoting in the normal version of the code.
